---
title: "Class Reference"
author: "Derek Willis"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the ~ is equivalent of /Users/[youruser] on a Mac).

```{r}
setwd("~/jour472/data_journalism_2024_fall")


```

### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)
```
### How to print in R

``` {r}
print ("Hello World")

```

### How to Add in R
``` {r}
2+2
```

### How to use variables in R
``` {r}
number <- 2
```
``` {r}
number * number
```
### How to install packages in R
```{r}
install.packages('tidyverse')
```
### What Obs. and Variables are in R
Obs is short for observation, which is another way of saying rows or records. Variables is another way of saying columns or fields
### How to preview your data in R
```{r}
glimpse(maryland_winred_contributions)
```
### How to preview your data in R (more organized)
```{r}
head(maryland_winred_contributions)
```
### How to decided what is included in your data set in R 
Note: From what I know right now, 'group-by()' decides what category is first, summarise includes the other categories included and arrange is how you sort your data (desc=decending)
```{r}
maryland_winred_contributions |>
  group_by(date) |>
  summarise(
    count_contribs = n()
  ) |>
  arrange(desc(count_contribs))
```