---
title: "Class Reference"
author: "Derek Willis"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the ~ is equivalent of /Users/[youruser] on a Mac).

```{r}
setwd("~/jour472/data_journalism_2024_fall")


```

### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)
```
### How to print in R

``` {r}
print ("Hello World")

```

### How to Add in R
``` {r}
2+2
```

### How to use variables in R
``` {r}
number <- 2
```
``` {r}
number * number
```
### How to install packages in R
```{r}
install.packages('tidyverse')
```
### What Obs. and Variables are in R
Obs is short for observation, which is another way of saying rows or records. Variables is another way of saying columns or fields
### How to preview your data in R
```{r}
glimpse(maryland_winred_contributions)
```
### How to preview your data in R (more organized)
```{r}
head(maryland_winred_contributions)
```
### How to decided what is included in your data set in R 
Note: From what I know right now, 'group-by()' decides what category is first, summarise includes the other categories included and arrange is how you sort your data (desc=decending)
```{r}
maryland_winred_contributions |>
  group_by(date) |>
  summarise(
    count_contribs = n()
  ) |>
  arrange(desc(count_contribs))
```
### How to do math and create new columns in R
```{r}
 mutate( percent_election_day = election_day/votes)
```
### Mutate is also useful for standardizing data - for example, making different spellings of, say, cities into a single one.
```{r}
standardized_maryland_cities <- maryland_cities |>
  mutate(
    upper_city = str_to_upper(city)
)
```
### How to decide the columns that you want to display
```{r}
primary_24 |>
  select(office_name, office_district, candidate_name, party, county_name, election_day, votes) |>
  mutate(
  percent_election_day = (election_day/votes)*100
)  |> 
  arrange(desc(percent_election_day))
```
### How to decide the columns that you want to display
```{r}
primary_24 |>
  select-(office_name, office_district, candidate_name, party, county_name, election_day, votes) |>
  mutate(
  percent_election_day = (election_day/votes)*100
)  |> 
  arrange(desc(percent_election_day))
```
### How to use `case_when` to give our new category column a value using some standard numeric logic.
```{r}
maryland_winred_categories <- maryland_winred_contributions |>
  mutate(
    amount_category = case_when(
        amount < 100 ~ "Less than $100",
        amount >= 100 & amount < 500 ~ "Between $100 and $499",
        amount >= 500 & amount < 1500 ~ "Between $500 and $1499",
        amount >= 1500 & amount < 3000 ~ "Between $500 and $2999",
        amount >= 3000 ~ "$3,000 or more"
      )
  )
head(maryland_winred_categories)
```
### How to clean up the column names
```{r}
cleaned_conowingo <- conowingo |>
  clean_names()
```
### How to remname column names
```{r}
cleaned_conowingo <- conowingo |>
  clean_names() |> 
  rename(linenumber = x1_linenumber)

# display the cleaned dataset
cleaned_conowingo
```
###fully clean data
```{r}
# cleaning function
cleaned_conowingo <- conowingo |>
  clean_names() |> 
  rename(linenumber = x1_linenumber) |> 
  mutate(amount = as.numeric(amount)) |> 
  distinct()
  

# display the cleaned dataset
cleaned_conowingo

```
pt 2
```{r}
# cleaning function
cleaned_conowingo <- conowingo |>
  clean_names() |> 
  rename(linenumber = x1_linenumber) |> 
  mutate(amount = as.numeric(amount)) |> 
  distinct() |>
  mutate(zip = str_sub(zip, start=1L, end=5L)) |>
  mutate(city = str_to_title(city))
  

# display the cleaned dataset
cleaned_conowingo

```
pt 3 
```{r}
# cleaning function
cleaned_conowingo <- conowingo |>
  clean_names() |> 
  rename(linenumber = x1_linenumber) |> 
  mutate(amount = as.numeric(amount)) |> 
  distinct() |>
  mutate(zip = str_sub(zip, start=1L, end=5L)) |>
  mutate(city = str_to_title(city)) |>
  mutate(city = case_when(
    str_detect(city,"^Conowing") ~ "Conowingo",
    TRUE ~ city
  ))
  

# display the cleaned dataset
cleaned_conowingo

```